A New Hybrid Machine Learning Method for Stellar Parameter Inference

### 2025-03-05 paper/paper2/paper.tex
GG?!

Ideally this process would be monolithic, with the end-to-end spectral inference code powered by a single autodiffable machine learning framework, like PyTorch \textbf{\citep{pytorch}} or JAX \citep{jax}.
However, here we separate the problem into three pieces, only the first of which is currently differentiable.
\textbf{In \autoref{sec:section2},} we scale out the \texttt{blase} method to 1,314 precomputed synthetic spectral model clones, yielding a downloadable archive of pretrained machine learning models with 128,723 unique spectral lines.
\textbf{In \autoref{sec:section3},} we \textbf{then} fit manifolds mapping stellar parameters to uniformly-derived spectral line parameters using regular grid linear interpolation.
Finally, \textbf{in \autoref{sec:section4},} we show how reconstructions of the spectra using said manifolds can be used for inferring stellar fundamental properties from spectra.
This final step resembles the aims of ``atmospheric retrievals'', but in principle should be faster, adaptive, and reusable.
An overview of this process is shown in \autoref{fig:figure1}.

@@

\section{Cloning the PHOENIX Model Grid}
\label{sec:section2}
\subsection{The PHOENIX Subset}
\textbf{For the purposes of this study, w}e chose the widely-adopted PHOENIX synthetic spectral model grid \citep{PHOENIX}.
Our approach can be straightforwardly applied to any other model grid in the future, including substellar atmosphere grids such as Sonora \citep{bobcat, cholla, diamondback, elfowl}, but for now we limit our scope to a \textbf{rectilinear} subset of the PHOENIX grid, focusing on near solar metallicities and a broad range of effective temperature and surface gravity, with details given in \autoref{tab:table1}.
\textbf{This approach is taken due to the computational cost of interpolation algorithms.
Future versions with more advanced models will be able to bypass the limitation of a rectilinear subset.}

@@

\subsection{Inference Algorithm}
\textbf{The goals of this study's inference algorithm are to infer the stellar parameters $T_\text{eff}$, $\log(g)$, and [Fe/H] by comparing a spectrum with our reconstructions and solving the optimization problem.}
We elected to use Bayesian optimization \textbf{for this component}, specifically the \texttt{gp\_minimize} function from the \texttt{scikit-optimize} library \citep{skopt}.
This algorithm uses a Gaussian Process to model the objective function, which in this case was the RMS (Root-Mean-Square) loss between the interpolated spectrum $\mathsf{M}$ and the true spectrum $\mathsf{D}$, defined as:
\begin{linenomath*}
\begin{gather}
    \mathcal{L} = \ev{(\mathsf{M} - \mathsf{D})^2}^{1/2}
\end{gather}
\end{linenomath*}
The optimizer was configured to first run 100 random evaluations to seed the surrogate model, then run 20 more evaluations now guided by the surrogate model.
\textbf{A random evaluation is simply a randomly generated tuple of stellar parameters within the parameter space of the PHOENIX generator, which is then evaluated with the generator to get a reconstruction and then compared with the inference target to retrieve its RMS error.
The surrogate model is the approximation to the true objective function that the optimizer constructs based on the set of random evaluations (in this case defined by a set of 100 points rather than a fully continuous, infinitely resolved function)
Guided evaluations are different, they are not randomly generated tuples of stellar parameters.
They use the surrogate model in order to select new test points for the minimum of the objective function.}
This totals to 120 evaluations, \textbf{100} to create \textbf{the} surrogate model, then \textbf{20} guided evaluation\textbf{s}, which was deemed sufficient for this study.
Fine-tuning these numbers is possible, but simply not warranted for a proof-of concept method.
\textbf{Tweaking these numbers could lead to a more accurate inference, however it would incur additional computational cost as well.}
One inference takes on average \textbf{200 seconds, using the computer with specifications in \autoref{tab:table3}}.

@@

\subsection{Bayesian Optimizer Performance}
To test the performance of the inference algorithm, we used the PHOENIX subset itself.
\textbf{This is possible for two reasons: first, the PHOENIX generator has not truly memorized the PHOENIX subset due to its Voigt profile basis and the ignoring of weaker features, and secondly, random samples within the parameter space are extremely unlikely to be PHOENIX grid points themselves, avoiding any overlap.}

@@

In \autoref{fig:figure8}, we show the residuals of the inferred stellar parameters with respect to the true values.
\begin{figure*}
    \centering
    \includegraphics[width=\textwidth]{figure8}
    \caption{\textbf{Residual plots for the GP minimization inference algorithm, all plotted with respect to $T_\text{eff}$ (values aggregated across $\log(g)$ and [Fe/H]).
    Notice the interesting change in behavior of all residuals at around $T_\text{eff} = 7000$ K.}}
    \label{fig:figure8}
\end{figure*}
\textbf{We see that there is a distinction between the algorithm's behavior above and below $T_{\mathrm{eff}} = 7000$ K, so we define low and high temperature regimes accordingly.
For $T_{\mathrm{eff}}$ < 7000 K, the inferred} $T_{\mathrm{eff}}$ differed from the true \textbf{value} by an average of \textbf{93} K, $\log(g)$ \textbf{0.24 dex}, and [Fe/H] \textbf{0.056} dex \textbf{(standard deviations of 139 K, 0.40 dex, and 0.15 dex, respectively)}.
\textbf{For $T_{\mathrm{eff}} \ge$ 7000 K, the inferred $T_{\mathrm{eff}}$ differed from the true value by an average of 347 K, $\log(g)$ 0.26 dex, and [Fe/H] 0.16 dex (standard deviations of 293 K, 0.32 dex, and 0.23 dex, respectively).}
From this, we can see that $T_{\mathrm{eff}}$ was the most accurately inferred parameter, followed by $\log(g)$, and then [Fe/H].

>>>>

Ideally this process would be monolithic, with the end-to-end spectral inference code powered by a single autodiff\textbf{erenti}able machine learning framework, like PyTorch \textbf{\citep{pytorch}} or JAX \citep{jax}.
However, here we separate the problem into three pieces, only the first \textbf{currently being auto}differentiable.
\textbf{In \autoref{sec:section2},} we scale out the \texttt{blase} method to 1,314 precomputed synthetic spectral model clones, yielding a downloadable archive of pretrained machine learning models with 128,723 unique spectral lines.
\textbf{In \autoref{sec:section3},} we \textbf{then} fit manifolds mapping stellar parameters to uniformly-derived spectral line parameters using regular grid linear interpolation.
Finally, \textbf{in \autoref{sec:section4},} we show how reconstructions of the spectra using said manifolds can be used for inferring stellar fundamental properties from spectra.
This final step resembles the aims of ``atmospheric retrievals'', but in principle should be faster, adaptive, and reusable.
An overview of this process is shown in \autoref{fig:figure1}.

@@

\section{Cloning the PHOENIX Model Grid}
\label{sec:section2}
\subsection{The PHOENIX Subset}
\textbf{For the purposes of this study, w}e chose the widely-adopted PHOENIX synthetic spectral model grid \citep{PHOENIX}.
Our approach can be straightforwardly applied to any other model grid in the future, including substellar atmosphere grids such as Sonora \citep{bobcat, cholla, diamondback, elfowl}, but for now we limit our scope to a \textbf{rectilinear} subset of the PHOENIX grid, focusing on near solar metallicities and a broad range of effective temperature and surface gravity, with details given in \autoref{tab:table1}.
\textbf{This approach is taken due to the computational cost and storage impact of interpolation algorithms.
Future versions with more advanced models will be able to cover a broader scope and bypass the limitation of a rectilinear subset.}

@@

\subsection{\textbf{Gaussian Process Minimization}}
\textbf{The goals of this study's inference algorithm are to infer the stellar parameters $T_\text{eff}$, $\log(g)$, and [Fe/H] by comparing a spectrum with our reconstructions and solving the optimization problem.}
We elected to use Bayesian optimization \textbf{for this component}, specifically the \texttt{gp\_minimize} function from the \texttt{scikit-optimize} library \citep{skopt}.
This algorithm uses a Gaussian Process to model the objective function, which in this case was the RMS (Root-Mean-Square) loss between the interpolated spectrum $\mathsf{M}$ and the true spectrum $\mathsf{D}$, defined as:
\begin{linenomath*}
\begin{gather}
    \mathcal{L} = \ev{(\mathsf{M} - \mathsf{D})^2}^{1/2}
\end{gather}
\end{linenomath*}
The optimizer was configured to first run 100 random evaluations to seed the surrogate model, then run 20 more evaluations now guided by the surrogate model.
\textbf{A random evaluation is simply a randomly generated tuple of stellar parameters within the parameter space of the PHOENIX generator, which is then evaluated with the generator to get a reconstruction and then compared with the inference target to retrieve its RMS error.
The surrogate model is the approximation to the true objective function that the optimizer constructs based on the set of random evaluations (in this case defined by a set of 100 points rather than a fully continuous, infinitely resolved function)
Guided evaluations are different, they are not randomly generated tuples of stellar parameters.
They use the surrogate model in order to select new test points for the minimum of the objective function.}
This totals to 120 evaluations, \textbf{100} to create \textbf{the} surrogate model, then \textbf{20} guided evaluation\textbf{s}, which was deemed sufficient for this study.
Fine-tuning these numbers is possible, but simply not warranted for a proof-of concept method.
\textbf{Tweaking these numbers could lead to a more accurate inference, however it would incur additional computational cost as well.}
One inference takes on average \textbf{200 seconds, using the computer with specifications in \autoref{tab:table3}}.

@@

\subsection{\textbf{Inference Accuracy}}
To test the performance of the inference algorithm, we used the PHOENIX subset itself.
\textbf{This is possible for two reasons: first, the PHOENIX generator has not truly memorized the PHOENIX subset due to its Voigt profile basis and the ignoring of weaker features, and secondly, random samples within the parameter space are extremely unlikely to be PHOENIX grid points themselves, avoiding any overlap.}

@@

In \autoref{fig:figure8}, we show the residuals of the inferred stellar parameters with respect to the true values.
\begin{figure*}
    \centering
    \includegraphics[width=\textwidth]{figure8}
    \caption{\textbf{Residual plots for the GP minimization inference algorithm, all plotted with respect to $T_\text{eff}$ (values aggregated across $\log(g)$ and [Fe/H]).
    Notice the interesting change in behavior of all residuals at around $T_\text{eff} = 7000$ K.}}
    \label{fig:figure8}
\end{figure*}
\textbf{We see that there is a distinction between the algorithm's behavior above and below $T_{\mathrm{eff}} = 7000$ K, so we define low and high temperature regimes accordingly.
For $T_{\mathrm{eff}}$ < 7000 K, the inferred} $T_{\mathrm{eff}}$ differed from the true \textbf{value} by an average of \textbf{93} K, $\log(g)$ \textbf{0.24 dex}, and [Fe/H] \textbf{0.056} dex \textbf{(standard deviations of 139 K, 0.40 dex, and 0.15 dex, respectively)}.
\textbf{For $T_{\mathrm{eff}} \ge$ 7000 K, the inferred $T_{\mathrm{eff}}$ differed from the true value by an average of 347 K, $\log(g)$ 0.26 dex, and [Fe/H] 0.16 dex (standard deviations of 293 K, 0.32 dex, and 0.23 dex, respectively).
We see that \texttt{blase3D} is much more effective at predicting stellar parameters of lower temperature spectra, presumably due to the larger amount of spectral lines that can be detected.}

\textbf{In the low-temperature regime, our median $T_\text{eff}$ residual is about 10 K, about half that of MINESweeper \citep{minesweeper} and \texttt{astroNN} \citep{leung2019}, and about ten times larger than ASPCAP \citep{ASPCAP}.
Median $\log(g)$ and [Fe/H] residuals are 0 dex, which are comparable to all three aforementioned methods, however \texttt{blase3D} is overall less consistent as the standard deviation of the residuals are about an order of magnitude larger.
At high temperatures, the median $T_\text{eff}$ residual changes drastically to about -434 K, which is significantly worse than the other methods.
$\log(g)$ becomes a few times less accurate, and both $\log(g)$ and [Fe/H] have large standard deviations, all leading to large inconsistencies and worse performance in this temperature regime.
\texttt{blase3D} at this stage does not quite reach the level of contemporary methods that can deal with more realistic data and cover much larger parameter spaces, but the fact that, in its best case, it can perform to within an order of magnitude using a drastically simpler implementation, sparks promise for future improvements.}
